---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.height = 8,
  dpi = 320
)
```

# tidyde

Perform differential expression analysis in a tidy way. Currently in early development. Only `limma` is supported. 

## Installation

`tidyde` requires the development version of several packages, including `ggplot2`, to function correctly. You can install the required packages for this vignette with the following code:

```{r, eval=FALSE}
install.packages(c("devtools", "tidyverse", "rlang"))
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("edgeR")
library(devtools)
install_github("latlio/tidyde")
```

# Tidy Differential Expression analysis

`tidyde` is a toolkit for working with gene differential expression analysis in R. Currently, it includes a sample count data set, called `GSE60450_Lactation-GenewiseCounts`, and sample metadata, called `SampleInfo_Corrected`. 

```{r, warning = FALSE, message = FALSE}
library(tidyde)
library(readr)
library(purrr)
library(dplyr)
library(stringr)
library(edgeR)
library(ggplot2)
counts <- read_delim("data/GSE60450_Lactation-GenewiseCounts.txt",
                     delim = "\t")
meta <- read_delim("data/SampleInfo_Corrected.txt",
                   delim = "\t") %>%
  mutate(FileName = str_replace(FileName, "\\.", "-"))
```

`tidyde` is composed of wrapper functions of the `limma` package that perform various steps of a standard differential expression analysis in a tidy way.

```{r}
# Set up objects
meta <- read_delim("data/SampleInfo_Corrected.txt",
                   delim = "\t") %>%
  mutate(FileName = str_replace(FileName, "\\.", "-"))

my_design <- check_sample_names(counts, c(1,2), meta, FileName) %>%
  purrr::pluck("meta") %>%
  make_design_matrix(., c("Status"))

id <- as.character(counts$EntrezGeneID)

# Analysis
res <- check_sample_names(counts, c(1,2), meta, FileName) %>%
  purrr::pluck("mod_count") %>%
  filter_genes(., id, "edgeR") %>%
  make_voom(., my_design) %>%
  model_limma() %>%
  make_contrasts(design_matrix = my_design, Statuspregnant, Statusvirgin) %>%
  model_bayes()
```

`tidyde` includes `broom`-like methods for cleaning differential expression fit results, although it currently is not officially supported by `broom`. I'm actively working on that. The `tidy()` function in `broom` puts results into a tidy data frame.

```{r, message = FALSE}
tidy.marray.lm(res)
```

# Visualization for Meta-Analysis

`tidyde` includes functionality for working with results in `ggplot2`, including quick plots for common visualizations.

With tidy data, most data visualizations for meta-analyses are easy to build from the ground up. Nevertheless, `tidyde` has several quick plot functions to make the process easier. For example, `volcano_plot()` takes a tidied DE analysis and plots a volcano plot.

```{r}
tidy.marray.lm(res) %>%
  volcano_plot()
```

Because the results are still `ggplot2` objects, it's easy to make changes to the plot.

```{r}
tidy.marray.lm(res) %>%
  volcano_plot() + 
  theme_minimal()
```

`tidyde` currently supports boxplots of counts, PCA plots, and volcano plots.
